Quantifying unmet ICU beds in tertiary hospitals in urban India using Propensity

### 2022-11-24 manuscript.Rmd
24 nov

---
title: 'Quantifying unmet ICU beds in tertiary hospitals in urban India using Propensity
  Score Matching '
author: "Maria Eriksson"
output:
  html_document: default
  word_document: default
bibliography: bibliography.bib
csl: vancouver.csl
---

@@

Abbreviations
============
ACS - American college of surgeons
ER - Emergency room
HIC - High-income countries
LMIC - Low- and middle-income countries
PSM - propensity score matching
RCT - Randomized case-control study
TITCO -Towards Improved Trauma Care Outcomes 
TQIP - Trauma Care Improvement Program

@@

Background
----------

Methods
-------

@@

Trauma globally and in India
----------------------------
Trauma is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity @traumaLMIC. 4.8 million people die from trauma annually, 90% are in LMIC and 20% in India alone @onemilliondead. Despite having only 1% of the global vehicle fleet, a disproportionate 10% of global traffic casualties occur in India @indiantrauma. Half of trauma mortality in India occur in-hospital. There is no national trauma registry in India, but isolated studies have shown that in-hospital mortality within 30 days of hospitalization is higher than similarly injured patients in HIC hospitals @indiantrauma. Globally, an estimated two million lives could be saved annually, if the appropriate and timely trauma care provided by hospitals in many high-income countries were available and provided in low to-middle-income countries (LMICs) @indiantrauma.

@@

Mortality due to accidents in India amounted to 421 104 people in 2019, which is 31.5 per 100.000 inhabitants. The traumatic mechanisms of injury from accidents covered in this thesis are road accidents, falls, assault and burns and these make up about half of the accident mortality @nationalcrime.

Trauma is a main cause of mortality and morbidity to otherwise healthy and young individuals, and dispropotionately affect this demographic. Most casualties affected the age group 30-45 years, followed by the age bracket 18-30 years @nationalcrime. People in this demographic are also oftentimes the primary or sole breadwinners for their families. Therefore the detrimental impact of trauma go beyond mortality or substantial loss of disease-free years of life for the affected individuals, it also carries a considerable financial and social harm on their family members and communities @traumaLMIC @nationalcrime

A diverse mix of public, for-profit and not-for-profit private actors provide health services in India, where private providers dominate. According to a WHO 2022 report, India has 76 709 non-ventilator supported and 39 476 ventilator-supported ICU beds @ICUbeds. Because of the type and severity of injuries, ICU resources are often essential in care of trauma patients @titco.

@@

Methods to examine trauma care and ensure equivalence
-----------------------------------------------------
In order to improve and ensure equivalence of intra-hospital care, trauma quality improvement programs (TQIP) have been developed. The pilot and most notable example is the TQIP developed by the American College of Surgeons ACS. Since the implementation of the ACS pilot, TQIP has evolved to a broad term that include many kinds of processes to improve the quality of trauma care, such as mortality and morbidity conferences, using audit filters or comparing risk adjusted outcomes over time and between centres. The aim is to find discrepancies in treatment for patients presenting with similar conditions. TQIPs aim to ensure patients get equivalent care within and across hospitals, to benchmark performance of health care providers versus their peers and to improve trauma care by developing best practice guidelines @tqip.

TQIP often originate in high-income countries (HIC), they are either not applicable nor easily transferred to the hospital conditions of LMIC @frailty @gunshot  @onemilliondead. For trauma patients, there exist significant structural differences between HIC and LMIC. These differences are in factors such as injury mechanisms and severity, health care resources as well as patient make-up @frailty @indiantrauma @traumaLMIC @onemilliondead. These differences make a direct transferral of HIC TQIP models onto LMIC irrelevant or misleading and there is a need to develop TQIP assessment applicable to local conditions.

<!-- Ni vil båda flytta fokus från detta, och jag förstår inte riktigt varför det är med kopplingen är luddig. Är det för att säga: det finns TQIPs som ger en fingervisning om hur ICU ska dimensioneras, bestyckas, bemannas och bedrivas. Men det finns inga färdiga TQIPs som kan användas för att kvantifiera ICU-bristen i Indien och vi tror det är vanskligt att använda de befintliga eftersom de skapats under så olika förhållanden. Och därför måste vi börja från noll? -->

@@

WHO and local health care workers believe there is a shortage of ICU beds in Indian hospitals, and that the lives of many patients would have been salvageable if the appropriate ICU resources had been available @ICUbeds. The differences of in-hospital trauma mortality in Indian hospitals was double that of high-income countries, and part of this difference could be explained by a shortage of ICU beds in India @indiantrauma .

@@

Quantify the unmet need for ICU beds in tertiaty hospitals in urban India.

Apply statistical analysis to a large data set of Indian trauma patients and determine how many out of the patients that were not admitted to the ICU reasonably should have been admitted given their condition. Their numbers in proportion to the number of ICU-admitted patients is indicative of the relative shortage of ICU beds.

@@

@@

Setting
-------

<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->
16 000 patients admitted to four public university hospitals in urban India are included in the TITCO dataset @titco. Initially the data set included 16047 patients, but to improve anonymization 47 patients were deleted from the dataset as described in the chapter anonymization. The participating hospitals were King Edward Memorial hospital (KEM) and Lokmanya Tilak Municipal General Hospital (LTMGH), Mumbai, Jai Prakash Narayan Apex Trauma Center (JPNATC), All India Institute of Medical Sciences, New Delhi and The Institute of Post-Graduate Medical Education and Research and Seth Sukhlal Karnani Memorial Hospital (SSKM), Kolkata. When the data was collected all hospitals were tertiary hospitals. JPNATC had almost 180 beds in a dedicated trauma centre, KEM had no specific trauma ward, LTMGH was a public university hospital with a specific trauma ward with 14 beds and SSKM was a  public university hospital with no specific trauma ward.

The participating hospitals are all public entities, so a majority of patients came from low socioeconomic circumstances in their respective communities.

@@

A dedicated project officer involved in the data collection was on site five days per week, approximately eight hours a day. The project officers covered night, evening and morning shifts in a way so all shift variants were covered every month of collection. Project officers worked in the area where trauma patients arrived to the hospital and collected data for patients arriving during their shift through direct observation. If the officer observed an examination being performed but not documented in the patient records were allowed to ask health care staff for the result. If a trauma patient arrived when a project officer was not on site, they were identified from the log books of the nurses on staff in the emergency department and their data was collected from patient records.

@@

<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->
The eligibility criteria for TITCO were patients who presented with injuries from trauma at the emergency departments of the participating hospitals with a mechanism of road traffic, railway, fall, burn or assault. Patients who were dead on arrival were excluded, as were patients who only had isolated limb injuries.

@@

This study does not requlre ethical permission, it refers to data collected and shared openly.

@@

Anonymization
-------------
Hospital identification number
Hospital name has been replaced with a identification number, and the conversion key has not been saved so there is not possible to reverse the conversion. Despite this, a high degree of anonymization can not be expected because the hospitals do have some distinguishable features and these are reflected in the data collected. For example,  70 % of the railroad accidents in the data set is in one of the hospitals, and one of the hosapitals is in close proximity to a big railroad hub.

Patient identification numbers
Patient names were never recorded, patients were identified by a identification number in the original unencrypted data set. In the anonymization process the identification number has been replaced with another random identification number, and the conversion key has not been saved so there is not possible to reverse the conversion.

Date observations
Dates recorded in the TITCO-I data set were shifted using a random offset of somewhere between 0 and 5 weeks for each patient, but without losing any relevant information. The offset was designed to not dissipate details that could be of potential use for researchers such as day of week and rough indication of season. The internal relations between dates have been preserved so that the timing of all recordings and discharge in chronological relation to admission are all intact.

It is important to note that all patients have sequence numbers that corresponded to the chronological order in which they arrived at any of the participating hospitals, and that sequence number has been kept intact. Keeping this sequence number variable intact allows for many useful operations such as examinations of trends in the data, while it is not possible to undo the date randomization by the sequence number variable, such an equation system is still underdetermined. Unfortunately this means it is also not possible to reconstruct the patient volumes at any given hospital on any given day and from that rate of resource engagement. Nor is it possible to distinguish if patients who were made to wait for their first assessment were so because their injuries were deemed as less severe or if it was because health care staff was overwhelmed which is an important distinction.

Age
Very old patients are few, they have been grouped together to prevent identification via the age parameter.

In addition, 47 patient data were randomly deleted to prevent identification via the sequence number variable.

Text based input
1% of the text-based inputs in ICD code fields and examination or operation findings were randomly deleted to obstruct identification by this variable.

@@

 Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->
There are 193 possible observations recorded for each patient. A general presentation of these observations is in table_nums('tab_1')

@@

The quantitative data include recorded parameters such as vital parameters, lab results, GCS total score and cumulative time in surgery or ICU, if applicable. Measurement or assessment techniques are standardized.

@@

Text data are freehand examination results and operation categorizations, as well as assigned ICD codes. These data are vast and contain a lot of useful information, but also introduce a lot of challenges. There is a lack of standardization, user dependence, heterogenous grouping caused by using too general ICD codes, and datamining challenges for this type of data. If it may cause inaccuracies or practical challenges, it shall be omitted.

@@

<!-- Describe any efforts to address potential sources of bias -->
The TITCO dataset are observational data collected from a clinical environment and not an experimental setting.  Medical interventions, examinations and ICU admission are not randomized, they are a consequence of a patients medical condition. There exists a causal effect between the independent and the dependent variables: a patient is sent to the ICU precisely because of the severity of their condition.

There exist a potential bias in the routines, resources and patient makeup of the participating hospitals. Becasue they are all tertiary hospitals, about two thirds of the patients have been referred from other health care providers, meaning that there is a selection bias in the patient populations. Further, hospitals with dedicated trauma wards have a propensity to receive transferred patients or police escorted patients with worse condition and prognosis.

As stated in chapter Settings, most patients have low socioeconomic status. Low socioeconomic status is associated with higher level of malnutrition and comorbidities and would have a negtive effect on vitals and the capacity of patients to tolerate certain interventions. This circumstance doesn't affect the analysis and internal validity, but reduce the generalizability of the results. 
<!-- (ÅTERKOM: har inte råd att refereas?) -->

Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but did not, these particular selection biases are not expected to skew the outcome, and therefore does not have to be adressed.

Hospitals with trauma wards could have lower admission criteria to the ICU than hospitals with no trauma ward. Because examining admission criteria to the ICU is part of the analysis it may be necessary to examine and address this bias by comparing ICU admitted patients between hospitals. Propensity score matching can be used to identify differences in ICU admission criteria between hospitals and find patients that were admitted but need not have been. These patients are excluded from predictor selection to reduce this bias. This is part of the development and training of the PSM model.

@@

Study size 
----------

<!-- Explain how the study size was arrived at -->
The data of 16000 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings.

@@

(a) Describe all statistical methods, including those used to control
for confounding
-->
Propensity score matching (PSM) is useful method for identifying and assembling a control group from within a data set where there is causality between the independent and dependent variables, and a randomized case-control (RCT) study to offset or balance this bias is impractical, immoral or otherwise improper. Propensity score is a statistical variable, understood as the likelihood of a subject receiving treatment conditional on observations of relevant covariates of that subject before treatment. In RCTs it is known from study design, but it is generally unknown in observational data. The rationale for PSM is that while it is unknown, this variable can be fairly well estimated using logistic regression.

The theory being examined in this masters thesis is that there exist a subset of the TITCO dataset that did not go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention. The thesis is that these patients were not admitted to the ICU only for lack of such resources, not by a medical assessment.

Translated into this context this means that in this data set exist a selection bias, there is causality between the independent variables medical conditions and interventions and the dependent variable ICU admittance. Patients were admitted to the ICU as a consequence of their medical condition. Performing a randomized case-control study and randomly assigning or withholding a live-saving treatment would obviously be immoral, but we believe that in the TITCO data set exist patients who were not admitted into the ICU by volition but for scarcity of such resources. Such a group would be akin to an non-experimental control group and PSM can be used to identify and assemble it.  If such a subset can be identified by PSM, then it naturally follows that there exists shortage of ICU beds in tertiary hospitals in India. It also follows that the shortage can be quantified by the proportionality of this subset versus the subset that did receive ICU care. If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.

The propensity score model is developed using the Training Data and tested using the Validation Data. The workflow contain a series of steps and the work flow should understood as linear, but iterative. The output of each step is the input for later steps. Therefore the output of the last step reflect the cumulative error from unfit assumptions made the second and fourth step. The output is tested against Validation Data. By reverting back and adjusting assumptions in an iterative process, the model is continuously improved and the accuracy and validity is improved.

<!-- ÅTERKOM inte klart bara gissat -->

The first step of PSM is to randomly split the data set into two parts, 60% and 40%. The 60% part is a Training Part used to delvelop and train the PSM, and the 40% part is the Validation Part. This is done because the inherent bias of the PSM method described in chapter Bias. Training Data is used in steps 2-4 and the output is then validated using Validation Data.

@@

Third step is to build a standard multivariable logistic regression model with ICU care as the outcome and the predictors selected in the second step as independent variables. This model is used to calculate the standard variable propensity score for all patients. The propensity score determines the likelihood that a patient will receive ICU care, given their condition.

Fourth step is PSM. In PSM members of the subset that did not get ICU care but had similar predictors - and therefore similar propensity scores - as the subgroup that did get ICU care, are matched to patients who did go to the ICU. There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria the outcome could inadvertedly vastly under- or overestimate the  need of ICU beds remain.  
The same procedure is repeated on the Validation Data using the same predictors and the same matching criteria, and this should produce a cohort comparable to the one trom Training Data. The proportion of this subset versus the subset that did receive ICU care is also the proprtion of ICU bed shortage. If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.

(b) Describe any methods used to examine subgroups and interactions

@@

Missing Data
-------------
The data completion rate is presented in table_nums('tab_1'). As the base assumption, all quantitative variables collected for each patient are considered potential recordings. Data that are missing are assumed missing because care for a patient did not motivate the collection of the data. For example, if serum creatinine was not recorded, it is assumed that the collection of this variable was not relevant for care of the patient.

There are some variables that should reasonably have been collected and documented given the patient condition, but omission will not be cause of dismissal unless these variables are predictors.

All patients with incomplete observations for any predictors will be discarded. For relevant qualitative variables, missing data did not automatically warrant dismissal, if the missing data could be indirectly proven. For example if the quantitative yes/no for an examination like a CT scan was missing, it was deemed a 'yes' if there was a CT scan result recorded. If both were missing, the patient data was dismissed.
All vital parameters are considered mandatory, irregardless of patient condition so patients without data of heart rate, systolic blood pressure or respiration rate at any time of their hospital stay, have been discarded.

>>>>

---
title: 'Quantifying unmet ICU beds in tertiary hospitals in urban India using Propensity
  Score Matching '
author: "Maria Eriksson"
output:
  html_document: default
  word_document: default
  pdf_document: default
bibliography: bibliography.bib
csl: vancouver.csl
---

@@

Abbreviations
============
ACS - American college of surgeons

AUC - Area Under Curve

ER - Emergency room

GCS - Glasgow coma scale

HIC - High-income countries

ISS - Injury Severity score

LMIC - Low- and middle-income countries

NISS - New Injury Severity Score

PSM - propensity score matching

RCT - Randomized case-control study

TITCO -Towards Improved Trauma Care Outcomes

TQIP - Trauma Care Improvement Program

@@

Background
----------
Trauma is a big cuse of mortality and morbidity in India, and it is a leading cause of death among young people. It has long been suggested that ICU resources are insufficient and that this is one of the factors that lead to Indian in-hospital motality being higher than in high-income countries.

Methods
-------
Propensity score matcing shall be applied to the TITCO dataset containing 16000 trauma patient details, in an effort to identify patients who did not go to the ICU but met the same treatment inclusion criteria as the patients that did receive ICU treatment, and as such reasonably should have received ICU treatment.

@@

Trauma globally and in India
----------------------------
Trauma is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity @traumaLMIC. 4.8 million people die from trauma annually, 90% are in LMIC and 20% in India alone @onemilliondead. Despite having only 1% of the global vehicle fleet, a disproportionate 10% of global traffic casualties occur in India @indiantrauma. Half of trauma mortality in India occur in-hospital, but this rate varies greatly by region and mode of injury @indiantrauma @nationalcrime @titco . There is no national trauma registry in India, but isolated studies have shown that in-hospital mortality within 30 days of hospitalization is higher than similarly injured patients in HIC hospitals @indiantrauma. Globally, an estimated two million lives could be saved annually, if the appropriate and timely trauma care provided by hospitals in many high-income countries were available and provided in low to-middle-income countries (LMICs) @indiantrauma.

@@

Trauma is a main cause of mortality and morbidity to otherwise healthy and young individuals, and dispropotionately affect this demographic. Most casualties affected the age group 30-45 years, followed by the age bracket 18-30 years @nationalcrime. People in this demographic are also oftentimes the primary or sole breadwinners for their families. Therefore the detrimental impact of trauma goes beyond mortality or substantial loss of disease-free years of life for the affected individuals, it also carries a considerable financial and social harm on their family members and communities @traumaLMIC @nationalcrime.

A diverse mix of public, for-profit and not-for-profit private actors provide health services in India, where private providers dominate. Public sector make up 20% of health care expenditure, and this 20% make up 1% of GDP. Both the proportion of public sector spending and the proportion of GDP is among the lowest in the world @goodstats. According to a WHO 2022 report, India has 76 709 non-ventilator supported and 39 476 ventilator-supported ICU beds @ICUbeds, out of which 10% of intitutions and 30% of ICU beds are public and financially feasibile for many Indians @indianicu @goodstats. Public hospitals have more or less perpetual shortage of ICU beds. If it is the attending physicians assessment that ICU care is medically necessary a patient or their relatives may choose to go to a private hospital, if the transfer is medically advisable. For a majority of patients it is financially unfeasible to pay out of pocket for the specialized ambulance or private hospitalization, that cost approx 100 times the average per capita income @perpetual @indianicu. Because of the type and severity of injuries, ICU resources are often essential in care of trauma patients @titco.

@@

Methods to examine trauma care and ensure equivalence
-----------------------------------------------------
In order to improve and ensure equivalence of intra-hospital care, trauma quality improvement programs (TQIP) have been developed. The pilot and most notable example is the TQIP developed by the American College of Surgeons ACS. Since the implementation of the ACS pilot, TQIP has evolved to a broad term that include many kinds of processes to improve the quality of trauma care, such as mortality and morbidity conferences, using audit filters or comparing risk adjusted outcomes over time and between centres. The aim is to find discrepancies in treatment for patients presenting with similar conditions. This difference in in-hospital mortality is likely multi-factorial, and any factors have been proposed to explain it such as differences in triage and delay-of-care, as well as shortage of hospital resources including ICU beds @onemilliondead ,@traumaLMIC . 
TQIPs aim to ensure patients get equivalent care within and across hospitals, to benchmark performance of health care providers versus their peers and to improve trauma care by developing best practice guidelines @tqip.

TQIP often originate in high-income countries (HIC), they are neither applicable nor easily transferred to the hospital conditions of LMIC @frailty @gunshot  @onemilliondead. For trauma patients, there exist significant structural differences between HIC and LMIC. These differences are in factors such as injury mechanisms and severity, health care resources as well as patient make-up @frailty @indiantrauma @traumaLMIC @onemilliondead. These differences make a direct transferral of HIC TQIP models onto LMIC irrelevant or misleading and there is a need to develop TQIP assessment applicable to local conditions.

<!-- Ni vil båda flytta fokus från detta, och jag förstår inte riktigt varför det är med kopplingen är luddig. Är det för att säga: det finns TQIPs som ger en fingervisning om hur ICU ska dimensioneras, bestyckas, bemannas och bedrivas Men det finns inga färdiga TQIPs som kan användas för att kvantifiera ICU-bristen i Indien och vi tror det är vanskligt att använda de befintliga eftersom de skapats under så olika förhållanden. Och därför måste vi börja från noll? -->

@@

WHO and local health care workers believe there is a shortage of ICU beds in Indian hospitals, and that the lives of many patients would have been salvageable if the appropriate ICU resources had been available @ICUbeds. In-hospital trauma mortality in Indian hospitals was double that of high-income countries, and part of this difference could be explained by a shortage of ICU beds in India @indiantrauma.

@@

Quantify the unmet need for ICU beds in tertiary hospitals in urban India.

The theory being examined in this master's thesis is that there exist a subset of the TITCO dataset that did not go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to be admitted. The hypothesis is that these patients were not admitted to the ICU only for lack of ICU resources, not by a medical assessment. 
<!-- Such a group would be akin to an non-experimental control group that PSM can be used to identify and assemble.   -->

If these patients can be identified, that is indicative there exists shortage of ICU beds in tertiary hospitals in India. The size of this group versus the size of the group that did receive ICU care indicate the magnitude of the ICU bed shortage. If, for every patient that did receive ICU care, there was another patient with similar predictors that should hade received ICU care based on our prediction model but did not due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.

@@

<!-- 
Definitions
------------
An ICU bed is a tiltable hospital bed but included in the term is also staffing and auxiliary equipment. The equipment is to administer oxygen and to monitor vitals. Some ICU beds are equipped with ventilators and if applicable length of ventilation has been registered.

Staffing is one doctor in the ward, and 1:4 nurses per patient but there is not always strict adherence to staffing guidelines. 
If the ICU ward is 100% full, patients might receive ventilation outside the ICU ward. Though the care these patients receive could very well be considered ICU care, they are not registered as such in the TITCO registry.
-->

@@

Study Setting
-------------

<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->
Four public university hospitals in urban India are included in the TITCO dataset [@titco]. The participating hospitals were King Edward Memorial hospital (KEM) and Lokmanya Tilak Municipal General Hospital (LTMGH), Mumbai, Jai Prakash Narayan Apex Trauma Center (JPNATC), All India Institute of Medical Sciences, New Delhi and The Institute of Post-Graduate Medical Education and Research and Seth Sukhlal Karnani Memorial Hospital (SSKM), Kolkata. When the data was collected all hospitals were tertiary hospitals. JPNATC had almost 180 beds in a dedicated trauma centre, KEM had no specific trauma ward, LTMGH was a public university hospital with a specific trauma ward with 14 beds and SSKM was a  public university hospital with no specific trauma ward. They are all end-stage public tertiary hospials in their respective areas and there is no possibility to refer patients onward within the public health care system, and ambulance equipment and traffic conditions oftentime makes it medically inadvisable.

The participating hospitals are all public entities, and a majority of patients came from low socioeconomic circumstances in their respective communities.

The data collection was designed in such a way that differences between hospitals would be minimized, but there are some differences in work methods, patient flows and resources that are unavoidable. Triage and patient work flow differed between hospitals. ICU admission criteria were consistent, even if only one hospital had formalized ICU admission criteria. Patients triaged toward the ICU ward were not considered ICU patients, until they had received an ICU bed and had been transferred. Between triage and ICU admission is the resuscitation room. Here - quite simplified - severely injured patients were stabilized before ICU admission and patients with intermediate injuries were kept for observation until most of them were transferred to a ward within one or a few days. The ICU ward occupancy was perpetually at 100 % and so resuscitated patients had to wait for admission with few exceptions. Patients in the resuscitation room were not considered ICU patients even if they were receiving a comparable level of care as ICU patients - patients receiving ionotropic support or patients who were intubated or ventilated. In LTMGH there was no resuscitation room and such a distinction, all admitted trauma patients were immediately considered ICU patients. Therefore for the purpose of this analysis LTMGH patients were considered ICU patients only if they had received ICU care >24 hours.

A dedicated project officer involved in the data collection was on site five days per week, approximately eight hours a day. The project officers covered night, evening and morning shifts in a way so all shift variants were covered every month of collection. Project officers worked in the area where trauma patients arrived to the hospital and collected data for patients arriving during their shift through direct observation. If the officer observed an examination being performed but not documented in the patient records were allowed to ask health care staff for the result. The project officers were strict observers and instructed not to intervene and delay care. If a trauma patient arrived when a project officer was not on site, they were identified from the log books of the nurses on staff in the emergency department and their data was collected from patient records.

@@

@@

<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->
The eligibility criteria for TITCO were all patients who presented with injuries from trauma at the emergency departments of the participating hospitals with a mechanism of road traffic, railway, fall, burn or assault. Patients who were dead on arrival were excluded, as well as patients who died within

Patients who only had isolated limb injuries were excluded

@@

This study does not require ethical permission, it refers to data collected and shared openly.

@@

<!-- Anonymization -->
<!-- ------------- GÄLLER INTE DENNA STUDIE SLAKTA OM BEHÖVS-->
<!-- Hospital identification number -->

<!-- Hospital name has been replaced with a identification number, and the conversion key has not been saved so there is not possible to reverse the conversion. Despite this, a high degree of anonymization can not be expected because the hospitals do have some distinguishable features and these are reflected in the data collected. For example,  70 % of the railroad accidents in the data set is in one of the hospitals, and one of the hosapitals is in close proximity to a big railroad hub.  -->

<!-- Patient identification numbers -->

<!-- Patient names were never recorded, patients were identified by a identification number in the original unencrypted data set. In the anonymization process the identification number has been replaced with another random identification number, and the conversion key has not been saved so there is not possible to reverse the conversion.  -->

<!-- Date observations -->

<!-- Dates recorded in the TITCO-I data set were shifted using a random offset of somewhere between 0 and 5 weeks for each patient, but without losing any relevant information. The offset was designed to not dissipate details that could be of potential use for researchers such as day of week and rough indication of season. The internal relations between dates have been preserved so that the timing of all recordings and discharge in chronological relation to admission are all intact.  -->

<!-- It is important to note that all patients have sequence numbers that corresponded to the chronological order in which they arrived at any of the participating hospitals, and that sequence number has been kept intact. Keeping this sequence number variable intact allows for many useful operations such as examinations of trends in the data, while it is not possible to undo the date randomization by the sequence number variable, such an equation system is still underdetermined. Unfortunately this means it is also not possible to reconstruct the patient volumes at any given hospital on any given day and from that rate of resource engagement. Nor is it possible to distinguish if patients who were made to wait for their first assessment were so because their injuries were deemed as less severe or if it was because health care staff was overwhelmed which is an important distinction. -->

<!-- Age -->

<!-- Very old patients are few, they have been grouped together to prevent identification via the age parameter. -->

<!-- In addition, 47 patient data were randomly deleted to prevent identification via the sequence number variable. -->

<!-- Text-based input -->
<!-- 1% of the text-based inputs in ICD code fields and examination or operation findings were randomly deleted to obstruct identification by this variable.  -->

@@

<!--  Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->
There are 193 possible observations recorded for each patient. A general presentation of these observations is in table_nums('tab_1')

@@

The quantitative data include recorded parameters such as vital parameters, lab results, Glasgow coma scale (GCS) total score and cumulative time in surgery or ICU, if applicable. Measurement or assessment techniques are standardized.

@@

Text data are freehand examination results and operation categorizations, as well as assigned ICD codes. These data are vast and contain a lot of useful information, but also introduce a lot of challenges. These data were not considered for lack of standardization, user dependence, heterogenous grouping caused by using too general ICD codes, and datamining challenges for this type of data.

@@

<!-- Describe any efforts to address potential sources of bias -->
The TITCO dataset is observational data collected from a clinical environment and not an experimental setting.  Medical interventions, examinations and ICU admission are not randomized, they are a consequence of a patient's medical condition. There exists a causal effect between the independent and the dependent variables: a patient is sent to the ICU precisely because of the severity of their condition. There is also a selection bias in which parameters are recorded and which examinations are performed, introducing detection and observer bias.

There exist a bias in the routines, resources and patient makeup of the participating hospitals as described in chapter Settings. Because they are all tertiary hospitals, about two-thirds of the patients have been referred from other health care providers. This means that there is a selection bias in the patient populations. Further, hospitals with dedicated trauma wards have a propensity to receive more transferred or police-escorted patients.

As stated in chapter Study Settings, most patients have low socioeconomic status. This circumstance doesn't affect the analysis and internal validity, but reduces the generalizability of the results. Most notably, if no ICU bed is available, and it is the attending physicians assessment that it is medically necessary, a patient or their relatives may choose to go to a private hospital. This is economically unfeasible for most of the patients in this study. This bias could not be corrected for, because such patients are pooled in the heterogenous group Discharged Against Medical Advice from where it is not possible to positively identify which patients in this category went on to private care.

Biases like these are expected in data collected in a clinical setting, the analysis model i chosen to adress this. Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but did not, these particular selection biases are not expected to skew the outcome, and therefore do not have to be addressed.

Hospitals with trauma wards could have different admission criteria to the ICU than hospitals with no trauma ward. The same goes for all equipment any hospital has as well as the proficiency of their physicians. Because examining admission criteria to the ICU is part of the analysis, any bias from differences in hospital admission criteria must be minimized. This is why the general linearized model is trained using a training subset and validated using a test subset, as described in the chapter Statistical methods. If this proves insufficient and unsatisfactory it may be necessary to examine and address this bias by comparing ICU admitted patients between hospitals using the same PSM methodology, but this will go beyond the time frame and scope of this thesis.

@@

@@

(a) Describe all statistical methods, including those used to control
for confounding
-->
<!-- Propensity score matching (PSM) is useful method for identifying and assembling a control group from within a data set where there is causality between the independent and dependent variables, and a randomized case-control (RCT) study to offset or balance this bias is impractical, immoral or otherwise improper.  -->

The propensity score model is developed using the Training Data and tested using the Test Data. The workflow contains a series of steps. The output of each step is the input for later steps. Steps three and four that involve development and validation of a multivariable regression model have several standard methods to test and improve model validity. By reverting back and adjusting assumptions in an iterative process, the model is continuously improved and the accuracy and validity is increased.

The first step of this analysis is to randomly split the data set into two parts, 60% and 40%. The 60% part is Training Data used to develop the propensity score model, and the 40% part is the Test Part. This is done to detect overfitting of the model to the Training Data.
Training Data is used in steps 2-3 and the resulting model is then validated using Test Data.

@@

Third step is to build a multivariable logistic regression model with ICU care as the outcome and the predictors selected in the second step as independent variables. The prediction model is based on the Training Data and predictors are weighed so that the the model awards patients that did go to the ICU corresponding high propensity. This model can then be used to predict new patients propensity to go to the ICU based on their baseline predictor values. By saving 40% in the Validation data set we have a set of patients that is new to the model. This allows to test how well the model correctly predicted ICU patients by awarding them high propensity scores.

Fourth step is to apply this model on patients from the Test Data. Using the model, the standard variable propensity score is calculated for each patient in the Test Data set. The propensity score determines the likelihood that a patient will receive ICU care, given their condition. If the model is suitable, it will correctly identify patients that did go to the ICU by awarding them high propensity scores, but non-ICU patients that were similar to this cohort will also get high propensity scores.

<!-- 
fråga: vi är ju ute efter ge vissa non-ICU pat höga propensity scores för att sedan kunna identifiera och filtrera fram dem i propensity score matchingen steg 5. Men hur identifierar man skillnaden mellan en patient som förmodligen borde fått ICU, och en patient modellen felaktigt spottat ut?
 -->
One standard method to evaluate how well a classification model performs in making binary predictions is apply it on new data and perform an Area Under Curve (AUC) calculation. In this calculation, the rate of true and false positives and negatives are calculated for each cut-off, which produces a curve as below. If the model has a 100% prediction accuracy, both sensitivity, specificity and AUC are all maximized at 1, the red lines. If the model has just as much ability to predict the outcome as a cointoss then sensitivity, specificity and AUC are 0.5, the blue line. For the purposes of this thesis, neither of these outcomes are desirable, and any model will perform somewhere between these extremes.

If there is no preference between rate of false positives and negatives, Youden's index is the optimal cut-off in propensity score with the best sensitivity and specificity. Patients that have a propensity score higher than this optimal cut-off in propensity score are patients that according to the model should have received ICU treatment. This number of patients can then be put in relation to the patients who received ICU treatment to calculate the unmet need of ICU beds.

<!-- peta in plot roc_test -->
![Exempel ROC kurva](C:/Users/maria/Desktop/stash/test_ROC.png)
Fifth step is bootstrapping step three and four to produce a confidence interval for the predictions of unmet ICU meds. Bootstrapping is a method of randomly resampling from the same data set and produce an output with some spread, a mean, standard deviation and 95% confidence interval. This will produce several samplings from which to produce several general linearized models, several AOC curves, several Youden's cut-offs, several estimates of patients whose propensity scores are higher than this cut-off and from this several proportions of unmet ICU bed needs. Several estimates with a mean, standard deviation and 95% confidence interval within which the best estimate of the unmet need is.

Bootstrapping is preferred over propensity score matching because of the weaknesses of the propensity score matching method and the available data size. 
<!-- återkom jag gissar bara -->

<!-- Fifth step is PSM. In PSM members of the subset that did not get ICU care are matched to patients who did go to the ICU, matching is based on patient propensity scores. There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable. There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria and inadvertently vastly under- or overestimate the  need of ICU beds remain.   -->

<!-- (b) Describe any methods used to examine subgroups and interactions -->

@@

Missing Data
-------------
The data completion rate is presented in table_nums('tab_1').

Rstudio default behaviour when making a multivariable regression model is to drop all entries with NA for predictors. This reduces the sample size, power and ultimately quality of the analysis. For the purposes of this statistical analysis, missing data in qualitative varibles is consdidered its own category. For quantitative data any variable with more than 10%  missing values are converted to an integer categorical variable and missing values treated as its own category. For systolic blood pressure and respiration rate, the interval brackets of the Revised Trauma Score was used, and for heart rate and blood oxygen saturation the brackets of National Early Warning Score (NEWS) was used. NEWS take into account whether or not the patient was receiving oxygen at the time of the recording.

Zero values for variables where it is possible to have zero values were left as is, such as respiration rate and heart rate if pulse was also zero and vice versa. However, when zero values are implausible, like when either heart rate or pulse were zero but the other was not, the observation is discarded. Data that were questionable such as when GCS observations did not add up were discarded.

<!-- FRÅGA: jag vill droppa noll i SC, eftersom noll är orimligt. men då droppar jag ca 3000 observationer på en variabel som inte ens har en stjärna i sginifikans för GLM. Droppa sc som prediktor alltogether? yes droppa sc
-->

<!-- (d)
Cohort study: If applicable, explain how loss to follow-up was addressed
